Normalization by Evaluation for Sized Dependent Types

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] "out of LR" proofs

\begin{enumerate}
\item % into
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}

@@

\begin{enumerate}
\item % into
We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.
The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
weakened assumption
 $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
% For the third, assume
% $\xi' : \Gamma'' \leq \Gamma'$.
% We have to show that there exists a neutral normal form $m'$ such that
% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
% $\RNe {\Gamma''} n m$ and
% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
% The latter equation implies
% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
% by irrelevant size application to
% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
% thus we are done by choosing $m' : = m \ann{a_0}$.
%
\qed
\end{enumerate}
%
\end{caselist}
\end{proof*}
} % END LONGVERSION
\begin{corollary}[Fresh variable injection]
  \label{cor:fresh}
  If\/ $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
  $\RG \Gamma T A$
  then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
\end{corollary}
\begin{proof}
  Since $\RNE {\cext \Gamma T} {\ind 0} {\var \Gamma} {T \slift}$
  by Lemma~\ref{lem:freshread},
  the goal follows from Thm.~\ref{thm:into} part (\ref{it:into}).
  % Assume $\xi : \Gamma' \leq \cext \Gamma T$.
  % Note that for the length
  % $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
  % we have $\ind i = \ind 0 \xi$, thus
  % $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
  % Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
  % the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
\end{proof}

>>>>

\begin{enumerate}
\item % into
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.

\item % out of
The premise is $\LR \Gamma t {T'} f {A'}$,
which means that for all $\xi : \Gamma' \leq \Gamma$ and
$\LR {\Gamma'} u {U\xi} a A$ we have
$\LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}$.

To show $\Gamma \der t \doteq \tR\,\down {S'} f : {T'}$
we assume $\xi : \Gamma' \leq \Gamma$ and produce a normal form $v$
such that $\R {\Gamma'} {\down{S'} f} {\lambda v}$ and
$\Gamma' \der t\xi = \lambda v : T'\xi$.
Induction hypothesis (\ref{it:into}) on
$\RNE {\cext{\Gamma'}{U\xi}} {\ind 0} {\var{\Gamma'}} {U\xi\slift}$
gives us $\LR {\cext{\Gamma'}{U\xi}} {\ind 0} {U\xi\slift} a A$
with $a := \up R {\var{\Gamma'}}$.
Thus, we can instantiate the assumption $\LR \Gamma t {T'} f {A'}$
to obtain
$\LR {\cext{\Gamma'}{U\xi}} {(t\xi\slift)\,\ind 0} {T(\xi\slift,\ind 0)} {f\,a} {B[a]}$.
Now induction hypothesis (\ref{it:out}) yields a normal form $v$ such that
$\R {\cext{\Gamma'}{U\xi}} {\down {S[a]} (f\,a)} v$
and
$\cext{\Gamma'}{U\xi} \der (t\xi\slift)\,\ind 0 = v : T(\xi\slift,\ind 0)$.
Since $a$ is the semantic version of the last bound variable,
$\R {\Gamma'} {\down{S'} f} {\lambda v}$ follows by definition of reification.
For the final goal, we $\lambda$-abstract the definitional equality to
$\Gamma' \der \lambda.\, (t\xi\slift)\,\ind 0 = \lambda v : \piT{(U\xi)} {(T(\xi\slift,\ind 0))}$
and with $\eta$-equality and the substitution laws we get
$\Gamma' \der t\xi = \lambda v : (\piT U T)\xi$.
\end{enumerate}

@@

\item % into
We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.
The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
weakened assumption $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
% For the third, assume
% $\xi' : \Gamma'' \leq \Gamma'$.
% We have to show that there exists a neutral normal form $m'$ such that
% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
% $\RNe {\Gamma''} n m$ and
% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
% The latter equation implies
% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
% by irrelevant size application to
% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
% thus we are done by choosing $m' : = m \ann{a_0}$.
%

\item % out of
To show $\RD \Gamma t {\down{S'} f} {T'}$ we assume $\xi : \Gamma' \leq \Gamma$ and
produce a normal form $v$ such that $\R{\Gamma'} {\down{S'}f} {\lambda v}$ and
$\Gamma' \der t\xi = \lambda v : T'\xi$.
It is sufficient to show
$\R{\erext{\Gamma'}\Size} {\down{S[\var{\Gamma'}]}(f \ann{\var{\Gamma'}})} v$
and
$\erext{\Gamma'}\Size \der (t\xi\slift)\,\ind 0 = v : T(\xi\slift,\ind 0)$.
These goals, in turn, follow by induction hypothesis (\ref{it:out}) on
$\LR{\erext{\Gamma'}\Size}{(t\xi\slift)\,\ind 0}{T(\xi\slift,\ind 0)}{f \ann{\var{\Gamma'}}}{A[\var{\Gamma'}]}$ which is an instance of our assumption
$\LR \Gamma t {T'} f {A'}$.
\qed
\end{enumerate}
%
\end{caselist}
\end{proof*}
} % END LONGVERSION
\begin{corollary}[Fresh variable injection]
  \label{cor:fresh}
  If\/ $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
  $\RG \Gamma T A$
  then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
\end{corollary}
\begin{proof}
  Since $\RNE {\cext \Gamma T} {\ind 0} {\var \Gamma} {T \slift}$
  by Lemma~\ref{lem:freshread},
  the goal follows from Thm.~\ref{thm:into} part (\ref{it:into}).
  % Assume $\xi : \Gamma' \leq \cext \Gamma T$.
  % Note that for the length
  % $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
  % we have $\ind i = \ind 0 \xi$, thus
  % $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
  % Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
  % the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
\end{proof}
